Wine Quality by Sula Al Hatem
========================================================

```{r message=FALSE}

#Loaing packages 

library(ggplot2)
library(knitr)
library(dplyr)
library(gridExtra)
library(ggExtra)
library(psych)
library(Simpsons)
library(memisc)

```

```{r echo=FALSE, Load_the_Data}
WhiteWine <- read.csv('wineQualityWhites.csv')
```

##Abstract:

White wine is a wine whose color can be straw-yellow, yellow-green, or 
yellow-gold. It is produced by the alcoholic fermentation of the non-colored 
pulp of grapes, which may have a skin of any color.

##Introduction:

The dataset is related to white variants of the Portuguese "Vinho Verde" wine.

The inputs include objective tests (e.g. PH values) and the output is based on 
sensory data (median of at least 3 evaluations made by wine experts). 
Each expert graded the wine quality between 0 (very bad) and 10 (very excellent)

I will be analyzing this dateset to get an idea on how the quality of wine is 
affected by each of the variables and by combinations of theses variables.

I would be also interested to see how some of these features affect each other.

##Dataset structure and summary

Let's take a look at what type of variables we have here:

```{r echo=FALSE, Univariate_Plots}
str(WhiteWine)
```

Looks like we have all variables as numbers, even the quality is an integer, 
i thinks some categorical variables could help, i will create three:

- A factor duplicate of quality

- A Factor with four levels for quality rate:

 quality <= 10, 'Excellent'
 
 quality <= 8, 'Very Good'
 
 quality <= 6, 'Fair'
 
 quality <= 4, 'Very Bad'

- A Factor with four levels for sweetness of quality: 

 0-4  : 'Dry'
 
 4-12 : 'Medium dry'
 
 12-45: 'Medium'
 
 45-66: 'Sweet' 

```{r echo=FALSE, message=FALSE, warning=FALSE, creating_categorical_variables}

#creating the rate variable from the quality
WhiteWine$rate <- NA
WhiteWine$rate <- ifelse(WhiteWine$quality <= 10, 'Excellent', WhiteWine$rate)
WhiteWine$rate <- ifelse(WhiteWine$quality <= 8, 'Very Good', WhiteWine$rate)
WhiteWine$rate <- ifelse(WhiteWine$quality <= 6, 'Fair', WhiteWine$rate)
WhiteWine$rate <- ifelse(WhiteWine$quality <= 4, 'Very Bad', WhiteWine$rate)

WhiteWine$rate <- factor(WhiteWine$rate, 
                         levels = c("Very Bad", 
                                    "Fair", 
                                    "Very Good", 
                                    "Excellent"))

# Creating a new factor variable from quality
WhiteWine$q.fact <- factor(WhiteWine$quality)


# creating sweetness variable for the levels of residual sugar
WhiteWine$sweetness <- cut(WhiteWine$residual.sugar, 
                           c(0, 4, 12, 45, 66), 
                           labels = c('Dry', 'Medium dry', 'Medium', 'Sweet'))
```

Now let's take a look at the five numbers of every variable:

```{r echo=FALSE}
summary(WhiteWine)
```

Median quality is 5 while the mean quality is 6.

# Univariate Plots Section

Seeing the distribution of the variables in plots will give even a clearer idea.

```{r echo=FALSE, message=FALSE, warning=FALSE}

b1 <- ggplot(data=WhiteWine, aes(x=1, y=fixed.acidity)) +
  geom_boxplot(alpha = .5,color = 'red') +
  scale_y_continuous(lim = c(3,12), breaks = seq(3,12,1)) 

h1 <- ggplot(data=WhiteWine, aes(x=fixed.acidity)) +
  geom_histogram( fill = I('#5760AB'), color = I('black'),aes(y = ..density..))+
  geom_density(alpha=.2, color="#FF6666") +
  scale_x_continuous(limits = c(3,12), breaks = seq(3,12,1))

grid.arrange(b1, h1, widths = c(2, 2))
```

The fixed.acidity looks like having almost a normal distribution. 
(after removing the highest outlier) the density line shows the peak as 6.45

I'm very interested to look at the other variables in the dataset.

```{r echo=FALSE, message=FALSE, warning=FALSE}

b2 <- ggplot(data=WhiteWine, aes(x=1, y=volatile.acidity)) +
  geom_boxplot(alpha = .5,color = 'red') +
  scale_y_continuous(lim = c(0,0.85), breaks = seq(0,0.85,.1)) 

h2 <- ggplot(data=WhiteWine, aes(x=volatile.acidity)) +
  geom_histogram( fill = I('#5760AB'), color = I('black'),aes(y = ..density..))+
  geom_density(alpha=.2, color="#FF6666") +
  scale_x_continuous(limits = c(0,0.85), breaks = seq(0,0.85,.1))

grid.arrange(b2, h2, widths = c(2, 2))
```

The volatile.acidity is how much acetic acid in wine 
(too high of levels can lead to an unpleasant, vinegar taste)

The distribution looks slightly right skewed. with the highest values at 0.25

```{r echo=FALSE, message=FALSE, warning=FALSE}

b3 <- ggplot(data=WhiteWine, aes(x=1, y=citric.acid)) +
  geom_boxplot(alpha = .5,color = 'red') +
  scale_y_continuous(lim = c(0,.9), breaks = seq(0,.9,.1)) 

h3 <- ggplot(data=WhiteWine, aes(x=citric.acid)) +
  geom_histogram( fill = I('#5760AB'), color = I('black'),aes(y = ..density..))+
  geom_density(alpha=.2, color="#FF6666") +
  scale_x_continuous(limits = c(0,.9), breaks = seq(0,.9,.1))

grid.arrange(b3, h3, widths = c(2, 2))
```

Small quantities of citric.acid can add 'freshness' and flavor to wines. 
The mode in this dataset is 0.3. there is a peak at 0.5 that might affect the 
taste and quality of the wine.

```{r echo=FALSE, message=FALSE, warning=FALSE}

b4 <- ggplot(data=WhiteWine, aes(x=1, y=residual.sugar)) +
  geom_boxplot(alpha = .5,color = 'red') +
  scale_y_continuous(lim = c(0,24), breaks = seq(0,24,5)) 

h4 <- ggplot(data=WhiteWine, aes(x=residual.sugar)) +
  geom_histogram( fill = I('#5760AB'), color = I('black'),aes(y = ..density..))+
  geom_density(alpha=.2, color="#FF6666") +
  scale_x_continuous(limits = c(0,24), breaks = seq(0,24,5))

grid.arrange(b4, h4, widths = c(2, 2))
```

The residual sugar is the amount of sugar remains after fermentation ends.
Based on that, we can tell if the wine is sweet or dry. The plot is right skewed
the peaks are at 1g and 2g with low residual sugar. 

I would like to see how is the distribution based on the sweetness levels:

```{r echo=FALSE, message=FALSE, warning=FALSE}
qplot(x = sweetness, data = WhiteWine, 
      fill = I('#5760AB'), color = I('black'), stat = 'count')

```

This is interesting, most of the wine in the dataset is "Dry" and "Medium Dry"
only a very small amount is "Sweet"

```{r echo=FALSE, message=FALSE, warning=FALSE}

b5 <- ggplot(data=WhiteWine, aes(x=1, y=chlorides)) +
  geom_boxplot(alpha = .5,color = 'red') +
  scale_y_continuous(lim = c(0,.22), breaks = seq(0,.22,0.05)) 

h5 <- ggplot(data=WhiteWine, aes(x=chlorides)) +
  geom_histogram( fill = I('#5760AB'), color = I('black'),aes(y = ..density..))+
  geom_density(alpha=.2, color="#FF6666") +
  scale_x_continuous(limits = c(0,.22), breaks = seq(0,.22,0.05))

grid.arrange(b5, h5, widths = c(2, 2))
```

The chlorides is the amount of salt in wine. The higher frequency here  is at 
around 0.04. the highest is 0.20. Too much salt will risk the quality

```{r echo=FALSE, message=FALSE, warning=FALSE}

b6 <- ggplot(data=WhiteWine, aes(x=1, y=free.sulfur.dioxide)) +
  geom_boxplot(alpha = .5,color = 'red') +
  scale_y_continuous(lim = c(0,120), breaks = seq(0,120,25)) 

h6 <- ggplot(data=WhiteWine, aes(x=free.sulfur.dioxide)) +
  geom_histogram( fill = I('#5760AB'), color = I('black'),aes(y = ..density..))+
  geom_density(alpha=.2, color="#FF6666") +
  scale_x_continuous(limits = c(0,120), breaks = seq(0,120,25))

grid.arrange(b6, h6, widths = c(2, 2))
```

A beautiful almost normal distribution for the free.sulfur.dioxide, there is 
a very noticeable peak at 26. free.sulfur.dioxide prevents microbial growth and 
the oxidation of wine

```{r echo=FALSE, message=FALSE, warning=FALSE}

b7 <- ggplot(data=WhiteWine, aes(x=1, y=total.sulfur.dioxide)) +
  geom_boxplot(alpha = .5,color = 'red') +
  scale_y_continuous(lim = c(0,310), breaks = seq(0,310,50)) 

h7 <- ggplot(data=WhiteWine, aes(x=total.sulfur.dioxide)) +
  geom_histogram( fill = I('#5760AB'), color = I('black'),aes(y = ..density..))+
  geom_density(alpha=.2, color="#FF6666") +
  scale_x_continuous(limits = c(0,310), breaks = seq(0,310,50))

grid.arrange(b7, h7, widths = c(2, 2))
```

This looks like  a normal distribution to me, with the peak is at around 125

```{r echo=FALSE, message=FALSE, warning=FALSE}

b8 <- ggplot(data=WhiteWine, aes(x=1, y=density)) +
  geom_boxplot(alpha = .5,color = 'red') +
  scale_y_continuous(lim = c(0.98,1.01)) 

h8 <- ggplot(data=WhiteWine, aes(x=density)) +
  geom_histogram( fill = I('#5760AB'), color = I('black'),aes(y = ..density..))+
  geom_density(alpha=.2, color="#FF6666") +
  scale_x_continuous(limits = c(0.98,1.01), breaks = seq(0.98,1.01,0.01))

grid.arrange(b8, h8, widths = c(2, 2))
```

The density depends on the percent of alcohol and sugar cont, there few high
frequency peaks in addition to the mode

```{r echo=FALSE, message=FALSE, warning=FALSE}

b9 <- ggplot(data=WhiteWine, aes(x=1, y=pH)) +
  geom_boxplot(alpha = .5,color = 'red') 
   
h9 <- ggplot(data=WhiteWine, aes(x=pH)) +
  geom_histogram( fill = I('#5760AB'), color = I('black'),aes(y = ..density..))+
  geom_density(alpha=.2, color="#FF6666")

grid.arrange(b9, h9, widths = c(2, 2))
```

pH has a beautiful normal distribution, without even removing any outliers.

It's interesting that pH levels are from 2.7 to 3.9, which is a characteristic of
white wine.

```{r echo=FALSE, message=FALSE, warning=FALSE}

b10 <- ggplot(data=WhiteWine, aes(x=1, y=sulphates)) +
  geom_boxplot(alpha = .5,color = 'red') +
  scale_y_continuous(lim = c(0.2,1), breaks = seq(0.2,1,0.1)) 

h10 <- ggplot(data=WhiteWine, aes(x=sulphates)) +
  geom_histogram( fill = I('#5760AB'), color = I('black'),aes(y = ..density..))+
  geom_density(alpha=.2, color="#FF6666") +
  scale_x_continuous(limits = c(0.2,1), breaks = seq(0.2,1,0.1))

grid.arrange(b10, h10, widths = c(2, 2))
```

The sulphate has a strange distribution distribution. There are some low amounts
at around 0.41 and 0.54 causing multiple peaks in the distribution, the highest 
is at 0.47

```{r echo=FALSE, message=FALSE, warning=FALSE}

b11 <- ggplot(data=WhiteWine, aes(x=1, y=alcohol)) +
  geom_boxplot(alpha = .5,color = 'red') 

h11 <- ggplot(data=WhiteWine, aes(x=alcohol)) +
  geom_histogram( fill = I('#5760AB'), color = I('black'),aes(y = ..density..))+
  geom_density(alpha=.2, color="#FF6666")

grid.arrange(b11, h11, widths = c(2, 2))
```

The alcohol level has an interesting distribution.There are no outliers. 
The dominating amount is around 9.1.There are many other high frequency values.
It almost looks like a missing rectangle shape with all of these peaks. 

After looking at all the input variables, I'm curious to see how the distribution
of quality from 0-9 and the distribution per rate look like:

```{r echo=FALSE, message=FALSE, warning=FALSE}

h12 <- ggplot(data = WhiteWine, 
       aes(x = q.fact)) +
  geom_histogram(fill = '#5760AB', color = 'black', 
                 stat = 'count')

  
h13 <- ggplot(data = WhiteWine, 
       aes(x = rate)) +
  geom_histogram(fill = '#5760AB', color = 'black', stat = 'count')

grid.arrange(h12, h13, widths = c(2, 2))

```

Very interesting: the higher frequency for quality is at 6, 5 is next. Looking 
at the 'rate' shows the dominant rate in the dataset is 'Fair' then 'Very Good'.

# Univariate Analysis

### The structure of the dataset

The dataset contains only physicochemical (inputs) and sensory (the output) 
variables with 4898 observation.

It came originally with 12 variable. I added two categorical variables to give 
a sense of the data. I also added a factorial duplicate of quality.

### Feature(s) of the dataset

The main focus would be to see which of the individual features affect quality
most.

I believe alcohol levels has an impact on the taste and quality of wine.

The level of pH, citric acid and the other acidic properties have a big impact 
on the taste and therefore the quality of wine. 

I am very interested to see how the the sweetness (residual sugar) contribute 
to the quality rating. i would think it's a matter of taste and it differs
per person, but I'm interested to take a look.

### Distribution and outliers:

From a first glance at the individual variables. We can see that most of the wine
samples are of 'Fair' quality (5-6), the second main quality is 'Very Good' (7-8)

- There are many outliers in the dataset.
- Alcohol and sulphate, citric.acid have strange distributions with many peaks
- pH, density, total.sulfur.dioxide and fixed acidity have normal distributions
- free.sulfur.dioxide, chloride, volatile.acidity and residual sugar have 
right skewed distribution

# Bivariate Plots Section

First, i will need to have an understanding of all the pair relationships 
between the quantitative variables.

```{r echo=FALSE, message=FALSE, warning=FALSE, Bivariate_Plots}
#i'm execludig the factorial variables 
pairs.panels(subset(WhiteWine, select = -c(X, q.fact, rate, sweetness)), 
             stars=TRUE, cex = 2)
```

I think a correlation table will also help:

```{r echo=FALSE, message=FALSE, warning=FALSE}
# i need to execlude the categorical variables and the index
cor(subset(WhiteWine, select = -c(X, q.fact, rate, sweetness)))
```

The plot and table are interesting, they confirm some of my thoughts about the 
relationships between the variables.

- There is a medium negative correlation between alcohol and density. Same with 
residual sugar, which makes sense since the later two are strongly correlated.

- pH has weak positive correlations with total.sulfur.dioxide, sulphates, alcohol
and quality

- quality is positively correlated with pH, sulphates, alcohol and 
free.sulfur.dioxide

- chloride, citric.acid, residual.sugar has a negative correlation with quality

- I think putting this in individual plots will help more understand this:

```{r echo=FALSE, message=FALSE, warning=FALSE}
ggplot(data = WhiteWine, aes(x = quality, y = alcohol)) +
  geom_jitter(alpha = .2, color = 'orange') +
  #geom_boxplot(alpha = .05, color = 'blue') +
    stat_summary(fun.y = "mean", geom = "point", 
               col = "red", shape = 18, size = 4) +
  stat_smooth(method = 'lm', fun.y = 'median', color = 'blue')
```

The positive correlation between alcohol and quality is obvious in this plot. 
yey to more alcohol in wine!

```{r echo=FALSE, message=FALSE, warning=FALSE}
ggplot(data = WhiteWine, aes(x = quality, y = citric.acid)) +
  geom_jitter(alpha = .1, color = 'orange') +
  #geom_boxplot(alpha = .05, color = 'blue') +
  stat_summary(fun.y = "mean", geom = "point", 
               col = "red", shape = 18, size = 4) +
  stat_smooth(method = 'lm', fun.y = 'median', color = 'blue') 
```

Surprisingly, citric acid doesn't look like having much effect on quality. It 
actually has a negative very week correlation of -0.009 which is unexpected for
me. Maybe it's because acidity makes the wine taste lighter and less rich and 
round

```{r echo=FALSE, message=FALSE, warning=FALSE}
ggplot(data = WhiteWine, aes(x = quality, y = residual.sugar)) +
  geom_jitter(alpha = .1, color = 'orange') +
  #geom_boxplot(alpha = .05, color = 'blue') +
  stat_summary(fun.y = "mean", geom = "point", 
               col = "red", shape = 18, size = 4) +
  stat_smooth(method = 'lm', fun.y = 'median', color = 'blue')
```

The quality if wine has a negative correlation with the amount of residual sugar
in it.

```{r echo=FALSE, message=FALSE, warning=FALSE}
ggplot(data = WhiteWine, aes(x = quality, y = density)) +
  geom_jitter(alpha = .1, color = 'orange') +
  #geom_boxplot(alpha = .05, color = 'blue') +
  stat_summary(fun.y = "mean", geom = "point", 
               col = "red", shape = 18, size = 4) +
  stat_smooth(method = 'lm', fun.y = 'median', color = 'blue')
```

Quality decreases with the increase of alcohol density, which makes sense, since
density and residual sugar are strongly correlated

```{r echo=FALSE, message=FALSE, warning=FALSE}
ggplot(data = WhiteWine, aes(x = quality, y = chlorides)) +
  geom_jitter(alpha = .05, color = 'orange') +
  #geom_boxplot(alpha = .05, color = 'blue') +
  stat_summary(fun.y = "mean", geom = "point", 
               col = "red", shape = 18, size = 4) +
  stat_smooth(method = 'lm', fun.y = 'median', color = 'blue')
```

Not surprisingly, there is a negative correlation between the quality of 
wine and the level of salt in it (chlorides). 

```{r echo=FALSE, message=FALSE, warning=FALSE}
ggplot(data = WhiteWine, aes(x = quality, y = volatile.acidity)) +
  geom_jitter(alpha = .05, color = 'orange') +
  #geom_boxplot(alpha = .05, color = 'blue') +
  stat_summary(fun.y = "mean", geom = "point", 
               col = "red", shape = 18, size = 4) +
  stat_smooth(method = 'lm', fun.y = 'median', color = 'blue')
```

There is a negative correlation between volatile acidity and quality of wine.
This makes sense as high amounts results in an unpleasant vinegar taste so it
decrease quality of wine.

```{r echo=FALSE, message=FALSE, warning=FALSE}
ggplot(data = WhiteWine, aes(x = quality, y = free.sulfur.dioxide)) +
  geom_jitter(alpha = .05, color = 'orange') +
  #geom_boxplot(alpha = .05, color = 'blue') +
  stat_summary(fun.y = "mean", geom = "point", 
               col = "red", shape = 18, size = 4) +
  stat_smooth(method = 'lm', fun.y = 'median', color = 'blue') 
```

There is slight increase in free.sulfur.dioxide with the increase in quality.

```{r echo=FALSE, message=FALSE, warning=FALSE}
ggplot(data = WhiteWine, aes(x = quality, y = sulphates)) +
  geom_jitter(alpha = .05, color = 'orange') +
  #geom_boxplot(alpha = .05, color = 'blue') +
  stat_summary(fun.y = "mean", geom = "point",
               col = "red", shape = 18, size = 4) +
  stat_smooth(method = 'lm', fun.y = 'median', color = 'blue')

```

Sulphate is positively correlated with quality, this also makes sense, since
sulphate acts as an antioxidant

```{r echo=FALSE, message=FALSE, warning=FALSE}

ggplot(data = WhiteWine, aes(x = quality, y = pH)) +
  geom_jitter(alpha = .05, color = 'orange') +
  #geom_boxplot(alpha = .05, color = 'blue') +
  stat_summary(fun.y = "mean", geom = "point",
               col = "red", shape = 18, size = 4) +
  stat_smooth(method = 'lm', fun.y = 'median', color = 'blue')

```

pH levels have a positive correlation with the quality of wine, which confirm the
negative correlation between quality and citric.acid.

I would like to see some of he relationships between the variables.

Since alcohol is the strongest correlation with quality, I'm 
curious to see what relationships it has with some of he other variables: 

```{r echo=FALSE, message=FALSE, warning=FALSE}

p1 <- ggplot(data = WhiteWine, aes(x = residual.sugar, y = alcohol)) +
  geom_point(color = 'blue', alpha = .1) +
  scale_x_continuous(limits = c(0,25)) +
  geom_smooth(color = 'red', method = 'lm')

p2 <- ggplot(data = WhiteWine, aes(x = citric.acid, y = alcohol)) +
  geom_point(color = 'blue', alpha = .1) +
  scale_x_continuous(limits = c(0,1)) +
  geom_smooth(color = 'red', method = 'lm')

p3 <- ggplot(data = WhiteWine, aes(x = fixed.acidity, y = alcohol)) +
  geom_point(color = 'blue', alpha = 0.1) +
  scale_x_continuous(limits = c(5, 11)) +
  geom_smooth(color = 'red', method = 'lm')

p4 <- ggplot(data = WhiteWine, aes(x = volatile.acidity, y = alcohol)) +
  geom_point(color = 'blue', alpha = 0.1) +
  scale_x_continuous(limits = c(0, 0.9)) +
  geom_smooth(color = 'red', method = 'lm')

grid.arrange(p1, p2, p3, p4, widths = c(3, 3))

```

alcohol has a strong negative correlation with residual sugar, and has medium 
negative correlations with citric.acid and with fixed.acidity.

The relationship for volatile.acidity with alcohol is surprising, i thought it
will be decreasing by the increase of alcohol considering it's negative relation
with quality, but it shows a slight increase.

Let's see how the pH behave with other variables:

```{r echo=FALSE, message=FALSE, warning=FALSE}

p4 <- ggplot(data = WhiteWine, aes(y = pH, x = citric.acid)) +
  geom_point(color = 'blue', alpha = 0.1) +
  scale_x_continuous(limits = c(0, 1)) +
  geom_smooth(color = 'red', method = 'lm')

p5 <- ggplot(data = WhiteWine, aes(y = pH, x = free.sulfur.dioxide)) +
  geom_point(color = 'blue', alpha = 0.1) +
  scale_x_continuous(limits = c(0, 100)) +
  geom_smooth(color = 'red', method = 'lm')

p6 <- ggplot(data = WhiteWine, aes(y = pH, x = alcohol)) +
  geom_point(color = 'blue', alpha = 0.1) +
  scale_x_continuous(limits = c(8, 14)) +
  geom_smooth(color = 'red', method = 'lm')

p7 <- ggplot(data = WhiteWine, aes(y = pH, x = density)) +
  geom_point(color = 'blue', alpha = 0.1) +
  scale_x_continuous(limits = c(0.989, 1.004)) +
  geom_smooth(color = 'red', method = 'lm')


grid.arrange(p4, p5, p6, p7, ncol = 2)

```

This is great and supports our earlier findings. pH has a negative correlation 
with citric.acid and density, and has a positive relation with alcohol and 
free.sulfur.dioxide.


# Bivariate Analysis

### Quality of wine

The four variables with positive correlation with quality (weak to strong) are:

- free.sulfur.dioxide 
- pH  
- sulphates   
- alcohol

### Other relationships

alcohol has the strongest positive correlation with the quality of wine. second
is sulphate

density and residual sugar are strongly correlated. so is free.sulfur.dioxide and
total.sulfur.dioxide

quality is negatively correlated with citric.acid. this was surprising for me.

alcohol and pH are positively correlated and both have negative relation with 
residual sugar and citric.acid.


# Multivariate Plots Section

I think it makes sense to start by looking at the two strongest positive relation
with quality:

```{r echo=FALSE, message=FALSE, warning=FALSE, Multivariate_Plots}
ggplot(data = WhiteWine, 
       aes(x = alcohol, y = sulphates, color = q.fact)) +
  geom_point(size = 1, alpha = .5) +
  facet_wrap(~ rate, ncol =4) +
  scale_color_brewer(type = 'seq') +
  geom_smooth(method = 'lm', se = FALSE)
```

Looks like combining higher amount of alcohol with higher amount of sulphates 
increase the quality of wine.

What about pH and alcohol combination with quality?

```{r echo=FALSE, message=FALSE, warning=FALSE}
ggplot(data = WhiteWine, 
       aes(x = alcohol, y = pH, color = q.fact)) +
  geom_point(size = 1, alpha = .5) +
  facet_wrap(~ rate, ncol =4) +
  scale_color_brewer(type = 'seq') +
  geom_smooth(method = 'lm', se = FALSE)
```

High level of pH with more alcohol also increases the quality of wine.

I believe pH and sulphates should have the same relationship, let's check:

```{r echo=FALSE, message=FALSE, warning=FALSE}
ggplot(data = WhiteWine, 
       aes(x = sulphates, y = pH, color = q.fact)) +
  geom_point(size = 1, alpha = .5) +
  facet_wrap(~ rate, ncol =4) +
  scale_color_brewer(type = 'seq') +
  geom_smooth(method = 'lm', se = FALSE)
```

As expected, high sulphates amount combined with pH increase the quality of wine.

citric.acid relationship with quality was a surprise for me, but we confirmed it
has the same relationship with alcohol. I'm curios to see the effect of alcohol 
with volatile.acidity on quality.

```{r echo=FALSE, message=FALSE, warning=FALSE}
ggplot(data = WhiteWine, 
       aes(x = alcohol, y = volatile.acidity, color = q.fact)) +
  geom_point(size = 1, alpha = .5) +
  facet_wrap(~ rate, ncol =4) +
  scale_color_brewer(type = 'seq') +
  geom_smooth(method = 'lm', se = FALSE)
```

The more alcohol in wine and the less volatile.acidity, the better the quality.

How about we try with pH:

```{r echo=FALSE, message=FALSE, warning=FALSE}
ggplot(data = WhiteWine, 
       aes(x = pH, y = volatile.acidity, color = q.fact)) +
  geom_point(size = 1, alpha = .5) +
  facet_wrap(~ rate, ncol =4) +
  scale_color_brewer(type = 'seq') +
  #scale_x_continuous(limits = c(2.7,4)) +
  #scale_y_log10() +
  geom_smooth(method = 'lm', se = FALSE)
```

The same, quality increases with low volatile.acidity and higher pH

hmm, i wonder if replacing the pH with fixed.acidity will change that:

```{r echo=FALSE, message=FALSE, warning=FALSE}
ggplot(data = WhiteWine, 
       aes(x = fixed.acidity, y = volatile.acidity, color = q.fact)) +
  geom_point(size = 1, alpha = .5) +
  facet_wrap(~ rate, ncol =4) +
  scale_color_brewer(type = 'seq') +
  #scale_x_continuous(limits = c(2.7,4)) +
  #scale_y_log10() +
  geom_smooth(method = 'lm', se = FALSE)
```

Not really, there are no effect on quality here.

free.sulfur.dioxide has a week positive correlation with quality, let's add pH
to the combination:

```{r echo=FALSE, message=FALSE, warning=FALSE}
ggplot(data = WhiteWine, 
       aes(x = pH, y = free.sulfur.dioxide, color = q.fact)) +
  geom_point(size = 1, alpha = .5) +
  facet_wrap(~ rate, ncol =4) +
  scale_color_brewer(type = 'seq') +
  #scale_x_continuous(limits = c(2.7,4)) +
  #scale_y_log10() +
  geom_smooth(method = 'lm', se = FALSE)
```

Higher free.sulfur.dioxide and higher alcohol means higher wine quality.

I would like to see how some relationships looks like when adding sweetness
categories:

```{r echo=FALSE, message=FALSE, warning=FALSE}
ggplot(data = WhiteWine, 
       aes(x = alcohol, y = density, color = sweetness)) +
  geom_point(size = 1, alpha = .5) +
  facet_wrap(~ rate, ncol =4) +
  scale_color_brewer(type = 'seq') +
  geom_smooth(method = 'lm', se = FALSE)
```

Low density and low sweetness with high alcohol levels make a better quality wine

```{r echo=FALSE, message=FALSE, warning=FALSE}
ggplot(data = WhiteWine, 
       aes(x = alcohol, y = pH, color = sweetness)) +
  geom_point(size = 1, alpha = .5) +
  facet_wrap(~ rate, ncol =4) +
  scale_color_brewer(type = 'div') +
  geom_smooth(method = 'lm', se = FALSE)
```

Sweet wine with low alcohol and low pH results is bad quality.

Dry and medium dry wine with specially higher alcohol level and higher PH 
increases the quality.

```{r echo=FALSE, message=FALSE, warning=FALSE}
ggplot(data = WhiteWine, 
       aes(x = alcohol, y = sulphates, color = sweetness)) +
  geom_point(size = 1, alpha = .5) +
  facet_wrap(~ rate, ncol =4) +
  scale_color_brewer(type = 'div') +
  geom_smooth(method = 'lm', se = FALSE)
```

same as before, sweet wine with low alcohol and low sulphates is bad quality.

Dry and medium dry wine with higher alcohol level and higher sulphates increases
the quality.

## Linear Model:

Let's fit our linear model with the features that has the strongest correlation
with quality.

```{r echo=FALSE, message=FALSE, warning=FALSE}
#let's first subset the data to remove the 

newdata = subset(WhiteWine, select = -c(X, q.fact, rate, sweetness))

```

### Simple Linear Model:
I'll be creating a simple linear model for the quality with the strongest 
relation; alcohol:

```{r echo=FALSE, message=FALSE, warning=FALSE, Linear_Model}
m1 <- lm(formula = quality ~ alcohol, data = newdata)

#checing the summary:
summary(m1)

```

- For every 8 mg of alcohol, the quality increases by 0.31

- We have a very small SE of 0.009 of the variation of alcohol/quality relation

- The p value is very significant, allows us to conclude a relationship between
alcohol and quality

- We have a 0.7973 residual SE, in which the change in quality will not follow 
a change in alcohol

The actual quality can deviate from the regression line by 0.797 points, so if 
we say that the average quality in the dataset is 2.582 and the residual SE is 
0.7973, the the percentage error (any prediction will still be off by) 30.879%

- 19% of the change in quality can be explained by the change in alcohol content

### Multiple Linear Model:
It would be a good idea to add more variables to the linear model:

```{r echo=FALSE, message=FALSE, warning=FALSE, Multiple_Linear_Model}

model <- lm(formula = quality ~ alcohol + sulphates + pH + free.sulfur.dioxide,
            data = newdata)

summary(model)

```

- The p value is very significant, allows us to conclude a relationship between
quality and all of the variables.

- We have a 0.788 residual SE, in which the change in quality will not follow 
a change in alcohol

The actual quality can deviate from the regression line by 0.7882 points, so if
we say that the average quality in the dataset is 1.303 and the residual SE is 
0.7882, the the percentage error (any prediction will still be off by) 60.49%

- 21% of the change in quality can be explained by the change in alcohol content

# Multivariate Analysis

### Relationships:

The strongest relationships are for alcohol, sulphates and pH. the combination 
of each two with quality. The higher the amount of these, the better the quality
of wine. 

The relationship for volatile.acidity with alcohol is surprising, i thought it
will be decreasing by the increase of alcohol considering it's negative relation
with quality, but it shows a slight increase. 

### Linear Model:

I created a simple linear model for alcohol vs. quality.
I also created a multiple linear model for the feature with the positive 
correlation with quality. 

The results confirm the relationship between the variables and quality.

------

# Final Plots and Summary

### Plot One
```{r echo=FALSE, message=FALSE, warning=FALSE, Plot_One}

h12 <- ggplot(data = WhiteWine, 
       aes(x = q.fact)) +
  geom_histogram(fill = '#5760AB', color = 'black', 
                 stat = 'count') +
  xlab("Quality Scale") 
  #ggtitle("Quality Distribution")
  

h13 <- ggplot(data = WhiteWine, 
       aes(x = rate)) +
  geom_histogram(fill = '#5760AB', color = 'black', stat = 'count') +
  xlab("Quality Rate") 
  #ggtitle("Rate Distribution")

grid.arrange(h12, h13, ncol = 2, 
             top= "Distribution of Wine Quality 
            By Scale (0 to 9) and by Rate (Very Bad to Excellent)")

```

### Description One
It's important to start by taking a look at the distribution of quality in the 
dataset. each entry is an average of three experts rating.
Most of the wine falls into the Fair category (5-6) the second ranking is for 
Very Good rate (7-8) 

It's good to keep this in mind, as no matter the correlation
with other variables, we are not reaching the level of excellent much.

### Plot Two
```{r echo=FALSE, message=FALSE, warning=FALSE, Plot_Two}
ggplot(data = WhiteWine, aes(x = quality, y = alcohol)) +
  geom_jitter(alpha = .2, color = 'orange') +
  #geom_boxplot(alpha = .05, color = 'blue') +
    stat_summary(fun.y = "mean", geom = "point", 
               col = "red", shape = 18, size = 4) +
  stat_smooth(method = 'lm', fun.y = 'median', color = 'blue') +
  xlab("Quality") +
  ylab("Alcohol content (% by Volume)") +
  ggtitle("Impact of Alcohol Content On Wine Quality")
```

### Description Two
Alcohol has the strongest positive relationship with quality. It means that better
quality wine has higher percentage of alcohol. although most of our dataset is 
in Fair or Very Good quality, as per our linear regression, alcohol contribute 
to around 19% to the change in wine quality.

### Plot Three

```{r echo=FALSE, message=FALSE, warning=FALSE, Plot_Three}
ggplot(data = WhiteWine, 
       aes(x = alcohol, y = sulphates, color = sweetness)) +
  geom_point(size = 1, alpha = .5) +
  facet_wrap(~ rate, ncol =4) +
  scale_color_brewer(type = 'div') +
  geom_smooth(method = 'lm', se = FALSE) +
  xlab("Alcohol Content (% by Volume)") +
  ylab("Sulphates Content (g/dm^3)") +
  ggtitle("Alcohol, Sulphates and Residual Sugar (Sweetness) Impact On Wine Qulity")
```

### Description Three
Alcohol and sulphates have the strongest relation with quality. The quality 
increases by the increase of each or both of them, it's worth looking at the 
relation of these three together and the sweetness of wine.

The quality increases by the increase of alcohol and the increase of sulphates,
while the sweetness decreases.

Dry and medium dry wine with higher alcohol level and higher sulphates increases
the quality.

------

# Reflection

The dataset contains only physicochemical and sensory data with 4898 wine sample
related to white variants of the Portuguese "Vinho Verde" wine.

At first, i wanted to get a sense of the data, so i started by checking the 
structure, summary of the data and creating plots to understand distribution of 
each individual variable.

I made some plots to more explore the visuals, and ended up exploring 
relationships between the quality of wine and many of the other variables. 
This helped me defining the strongest relationships in the dataset, which i
explored more with multiple variables plots.

I found out that the four variables with positive correlation with quality are:

- free.sulfur.dioxide 
- pH  
- sulphates   
- alcohol

I was curious to see the effect of other variables on alcohol, while i saw some
expected results, like the negative correlation with residual.sugar, citric.acid
and fixed.acidity, i was surprised with the positive relationship with 
volatile.acidity. Adding quality to the combination, turned the relation to a
negative one as expected, since quality has a strong positive correlation with
alcohol and a negative one with volatile.acidity.

I also looked at the relationship of pH with some of the variables, the results 
were as expected, 

I also created a simple linear model for quality with the strongest relation:
alcohol, this helped me figuring out the contribution of alcohol in the change
of quality.
I also created a multiple linear model for the main variables that positively 
affect h quality, this confirmed the relationship and showed their contribution
to the change in quality.

Some limitations was that i don't think this is complete dataset, as most of
wines are of Fair or Very Good Qualities (between 5 and 8 rating) I think with 
larger dataset with more experts to rate and more variety of wines, the dataset
would be more interesting to analyse, especially if e can have the individual
ratings included in the dataset rather than the average. It would also help if
the criteria for the rating is provided.
